---
---

## dplyr Functions

| Function    | Returns                                            |
|-------------+----------------------------------------------------|
| `filter`    | keep rows that satisfy conditions                  |
| `mutate`    | apply a transformation to existing [split] columns |
| `select`    | keep columns with matching names                   |
| `inner_join`| merge columns from separate tables into one table  |
| `group_by`  | split data into groups by an existing factor       |
| `summarize` | summarize across rows [and combine split groups]   |

The table above summarizes the most commonly used functions in
[dplyr](){:.rlib}, which we will demonstrate in turn on data from the U.S.
Census Bureau.
{:.notes}

===

### Filter

The `cbp` table includes character `NAICS` column. Of the 2 million
observations, lets see how many observations are left when we keep only the
2-digit NAICS codes, representing high-level sectors of the economy.

```{r lib, message = FALSE, handout = 0}
library(dplyr)
cbp2 <- filter(cbp,
  grepl('----', NAICS),
  !grepl('------', NAICS))
```

===

```{r}
str(cbp2)
```

Note that a logical "and" is implied when conditions are separated by commas.
(This is perhaps the main way in which `filter` differs from the base R `subset`
function.) Therefore, the example above is equivalent to `filter(grepl('----',
NAICS), !grepl('------', NAICS)`. A logical "or", on the other hand, must be
specified explicitly with the `|` operator.
{:.notes}

===

The [stringr](){:.rlib} package makes the use of pattern matching by [regular
expressions] a bit more maneageble, and streamlines this step.

```{r message = FALSE, handout = 0}
library(stringr)
cbp2 <- filter(cbp,
  str_detect(NAICS, '[0-9]{2}----'))
```

[regular expressions]: https://stringr.tidyverse.org/articles/regular-expressions.html

===

### Mutate

The `mutate` function is the [dplyr](){:.rlib} answer to updating or altering
your columns. It performs arbitrary operations on existing columns and appends
the result as a new column of the same length.

===

Here's one you've probably needed before:

```{r, handout = 0}
cbp3 <- mutate(cbp2,
  FIPS = str_c(FIPSTATE, FIPSCTY))
```

===

Multiple arguments to `mutate` produce multiple transformations.

```{r, handout = 0}
cbp3 <- mutate(cbp2,
  FIPS = str_c(FIPSTATE, FIPSCTY),
  NAICS = str_remove(NAICS, '-+'))
```

===

### Chaining Functions

All the functions from the [dplyr](){:.rpkg} package take a data frame as their
first argument, and they return a data frame. This consistent syntax is on
purpose. It is designed for easily chaining data transformations together:
creating a data pipeline that is easy to read and modify.

===

The "pipe" operator (`%>%`) takes the expression on its left-hand side and
inserts it, as the first argument, into the function on its right-hand side.
Equivalent to `sum(c(1,3,5))`, for example, we have:

```{r pipe_example_1}
c(1, 3, 5) %>% sum()
```

===

Additional arguments are accepted---the pipe only handles the first.

```{r pipe_example_2}
c(1, 3, 5, NA) %>% sum(na.rm = TRUE)
```

===

The pipe operator's main utility is to condense a chain of operations applied to
the same piece of data, when you don't want any intermediate results. We
can do the `filter` and `mutate` operations from above with one assignment.

===

```{r handout = 0, warning = FALSE}
cbp <- cbp %>%
  filter(
    str_detect(NAICS, '[0-9]{2}----')
  ) %>%
  mutate(
    FIPS = str_c(FIPSTATE, FIPSCTY),
    NAICS = str_remove(NAICS, '-+')
  )
```

===

### Select

To keep particular columns of a data frame (rather than filtering rows), use
the `select` function with arguments that match column names.

```{r}
names(cbp)
```

===

One way to "match" is by including complete names, each one you want to keep:

```{r eval = FALSE}
cbp %>%
  select(
    FIPS,
    NAICS,
    N1_4, N5_9, N10_19 # a better way?
  )
```

===

Alternatively, we can use a "[select helper](https://dplyr.tidyverse.org/reference/select.html#useful-functions)" to match patterns.

```{r select, handout = 0}
cbp <- cbp %>%
  select(
    FIPS,
    NAICS,
    starts_with('N')
  )
```
