---
excerpt: Wide or Long
---

## Gather

The [tidyr](){:.rlib} package's `gather` function reshapes "wide" data frames
into "long" ones.

```{r handout = 0}
library(tidyr)
tidy_trial <- gather(trial,
  key = "treatment",
  value = "response",
  -block)
```

All columns, accept for "block", are stacked in two columns: a "key" and a
"value". The key column gets the name `treatment` and the value column reveives
the name `response`. For each row in the result, the key is taken from the name
of the column and the value from the data in the column.
{:.notes}

===

```{r}
tidy_trial
```

Some notes on the syntax: a big advantage of [tidyr](){:.rlib} and
[dplyr](){:.rlib} is that each function takes a data frame as its first argument
and returns a new data frame. As we will see later, it makes it very easy to
chain these functions in a pipeline. All functions also use column names as
variables without subsetting them from a data frame (i.e. `block` instead of
`trial$block`).
{:.notes}

===

## Spread

Data can also fail to be tidy when a table is too long. The
Entity-Attribute-Value (EAV) structure common in large databases distributes
multible attributes of a single entity/observation into separate rows.

Remember that the exact state of "tidy" may depend on the analysis: the key is
knowing what counts as a complete observation. For example, the community
ecology package [vegan](){:.rlib} requires a matrix of species counts, where
rows correspond to species and columns to sites. This may seem like too "wide" a
format, but in the packages several multi-variate analyses, the abundance of a
species across multiple sites is considered a complete observation.
{:.notes}

===

Consider survey data on participant's age and income *stored* in a EAV
structure.

```{r, echo = FALSE}
survey <- read.delim(sep = ',', header = TRUE, text = "
participant,   attr, val
1          ,    age,  24
2          ,    age,  57
3          ,    age,  13
1          , income,  30
2          , income,  60
")
kable(survey)
```

===

Transform the data with the `spread` function, which "reverses" a `gather`.

```{r, handout = 0}
tidy_survey <- spread(survey,
  key = attr,
  value = val)
```

===

```{r}
tidy_survey
```


===

One difficulty with EAV tables is the nature of missing data; an entire row
rather than a single cell is missing. Think about what "missing data" could mean
here---perhaps you can supply a value instead of the `NA` in the previous
result.

```{r, handout = 0}
tidy_survey <- spread(survey,
  key = attr,
  value = val,
  fill = 0)
```

===

```{r}
tidy_survey
```
