---
excerpt: Wide or Long
---

## Pivot_longer

The [tidyr](){:.rlib} package can be used to reshape "wide" data frames into "long," tidy data frame. By making each observation a single row, it "lengthens" the data. 

We can use [tidyr](){:.rlib} `pivot_longer` function to reshape the table for drug trial data from above (named `trial`) into a tidy data frame.
{:.notes}

```{r handout = 0}
library(tidyr)
tidy_trial <- pivot_longer(trial,
  cols = c(drug, control, placebo),
  names_to = 'treatment',
  values_to = 'response')
```

All columns, except for "block", are stacked in two columns: a "name" and a
"value", as specified by the `cols` arguement. The name column gets the name `treatment` from the `names_to` argument,
and the `value` column receives the name `response` from the `values_to` argument. 
For each row in the result, the "name" (treatment) is taken from the name
of the column and the "value" (response) from the data in the column. The resulting table is saved as `tidy_trial`.
{:.notes}

```{r}
tidy_trial
```

Some notes on the syntax: a big advantage of [tidyr](){:.rlib} and
[dplyr](){:.rlib} is that each function takes a data frame as its first argument
and returns a new data frame. As we will see later, it makes it very easy to
chain these functions in a pipeline. All functions also use column names as
variables without subsetting them from a data frame (i.e. `block` instead of
`trial$block`).
{:.notes}

===

## Pivot_wider

Data can also fail to be tidy when a table is too long. The
Entity-Attribute-Value (EAV) structure common in large databases distributes
multiple attributes of a single entity/observation into separate rows.

Remember that the exact state of "tidy" may depend on the analysis: the key is
knowing what counts as a complete observation. For example, the community
ecology package [vegan](){:.rlib} requires a matrix of species counts, where
rows correspond to species and columns to sites. This may seem like too "wide" a
format, but in several multi-variate analyses, the abundance of a
species across multiple sites is considered a complete observation.
{:.notes}

===

Consider survey data on participant's age and income *stored* in a EAV
structure.

```{r, echo = FALSE}
survey <- read.delim(sep = ',', header = TRUE, strip.white = TRUE, text = "
participant,   attr, val
1          ,    age,  24
2          ,    age,  57
3          ,    age,  13
1          , income,  30
2          , income,  60
")
knitr::kable(survey)
```

===

Transform the data with the `pivot_wider` function, which "reverses" a `pivot_longer`. 

```{r, handout = 0}
tidy_survey <- pivot_wider(survey,
  names_from = attr,
  values_from = val)
```

The column names for the resulting data frame `tidy_survey` are taken from the `attr` column and the data from the `val` column.
{:.notes}

```{r}
tidy_survey
```

===

One difficulty with EAV tables is the nature of missing data; an entire row
rather than a single cell is missing. It is important to evaluate how to handle the "missing data." By default, "missing data" is coded as `NA`. 
{:.notes}

The `values_fill` argument allows you to specify values for this missing data. 

```{r, handout = 0}
tidy_survey <- pivot_wider(survey,
  names_from = attr,
  values_from = val,
  values_fill = 0)
```

```{r}
tidy_survey
```
