---
---

## Split-Apply-Combine

A very common data manipulation procedure is doing some "group-wise" operations
on a dataset and combing the results for each group into a single table. For
example, say you need to sum the number of employees *in each state* observed in
the health care sector?

===

## Grouping

The dplyr function `group_by` begins the process by indicating how the data frame should be split into subsets.

```{r summarize, handout = 0}
state_cbp_health_care <- cbp_health_care %>%
    group_by(FIPSTATE)
```

===

At this point, nothing has really changed:

```{r}
str(state_cbp_health_care)
```

The `group_by` statement does not change any values in the data frame; it only
adds attributes to the the original data frame. You can add multiple variables
(separated by commas) in `group_by`; each distinct combination of values across
these columns defines a different group.
{:.notes}

===

## Summarize

The operation to perform on each species is summing: we need to sum employment
are in each group.

```{r summarize_n, handout = 0}
state_cbp_health_care <- cbp_health_care %>%
    group_by(FIPSTATE) %>%
    summarize(EMP = sum(EMP))
```

===

```{r view_summarize}
str(state_cbp_health_care)
```

===

The "combine" part of "split-apply-combine" occurs automatically, when the attributes introduced by `group_by` are dropped. You can see attributes either by running the `str()` function on the data frame or by inspecting it in the RStudio *Environment* pane.

===

The function `n()` takes no arguments and returns the number of records in a group. Any function that collapses a vector input to a single output is a suitable function to use within `summarize`.

```{r handout = 0}
state_cbp_health_care <- cbp_health_care %>%
    group_by(FIPSTATE) %>%
    summarize(EMP = sum(EMP),
              counties = n())
```

===

But `group_by` can't read your mind, so be careful to type what you mean! 

```{r}
head(state_cbp_health_care)
```
