---
# haven't modified yet
---


## Putting it all together

The `mutate` function creates new columns from existing ones. The data frame returned has an additional column for each argument to `mutate`, unless a name is reused. Overwriting an existing column does not generate a warning.

===

The `count` variable just defined, for example, can be used to calculate the proportion of individuals represented by each species.

```{r mutate, handout = 0}
prop_1990_winter <- counts_1990_winter %>%
    mutate(prop = count / sum(count))
```

===

```{r}
head(prop_1990_winter)
```
For a concise way to apply the same transformation to multiple columns, check the `mutate_each` function. There is also a `summarize_each` function to perform the same aggregation operation on multiple columns.
{:.notes}

===

Both `mutate` and `summarize` can be used in the "apply" part of a "split-apply-combine" procedure. The difference is that the results are combine into data frames with differing numbers of rows.

===

Question
: How many rows do you expect in the result of a `mutate` operation?

Answer
: {:.fragment} The same number you started with.
