---
---

## Exercises

===

### Exercise 1

Now that we have a tidy form of `survey`, convert it to a `long_survey` data frame
using `gather`. The only difference between `survey` and `long_survey` should be
an additional row for zero income.

[View solution](#solution-1)
{:.notes}

===

### Exercise 2

Write code that returns the annual payroll data for the top level Construction
sector ("23----").

[View solution](#solution-2)
{:.notes}

===

### Exercise 3

Write code that fixes `state_cbp_health_care`, which should show the number
of counties and the total employment. Group the data using **both** `FIPSTATE`
and `FIPSCTY` and use the fact that a call to `summarize` only combines across
the lowest level of grouping.

[View solution](#solution-3)
{:.notes}

===
<!--

### Exercise 4

A "pivot table" is a transformation of tidy data into a wide summary table. First, data are summarized by *two* grouping factors, then one of these is "pivoted" into columns. Starting from the `animals` data frame, chain a `group_by` and `summarize` transformation into a [tidyr](){:.rlib} `spread` function to get the number of individuals counted in each month (as three columns) by species (as rows).

===
-->

## Solutions

===

### Solution 1

```{r}
gather(tidy_survey, key = "attr", value = "val", -participant)
```

[Return](#exercise-1)
{:.notes}

===

### Solution 2

```{r}
cbp_construction <- filter(cbp, NAICS == '23----')
select(cbp_construction, starts_with('FIPS'), starts_with('AP'))
```

[Return](#exercise-2)
{:.notes}

===

### Solution 3

```{r}
state_cbp_health_care <- cbp_health_care %>%
  group_by(FIPSTATE, FIPSCTY) %>%
  summarize(EMP = sum(EMP)) %>%
  summarize(EMP = sum(EMP), counties = n())
```

[Return](#exercise-3)
{:.notes}

<!--
===

### Solution 4

```{r}
group_by(animals, species_id, month) %>%
  summarize(count = n()) %>%
  spread(key = month, value = count, fill = 0)
```

[Return](#exercise-4)
{:.notes}
-->